---
title: "Phenoarch platform - Cleaning procedure - Point level using locfit"
author: "I.Sanchez"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Phenoarch platform - Cleaning procedure - Point level using locfit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!---
Program  : detectOutlierPoints.Rmd
Author   : I.Sanchez (MISTEA)
Objective: a vignette for phisStatR library
           require openSilexStatR library and flagPointLocfit() function
           The user has to adapt to his own analysis.
Creation : 25/05/2020
Update   : 28/05/2020
-->

***
# Objective 

Detection of outlier points in time courses of an experiment in PhenoArch greenhouse. This procedure can be used in any kind of time courses data sets. It uses a locfit smoothing function from the *locfit* library [2]. For each time course of a dataset, a locfit smoothing is applied, predictive confidence interval calculated (Y$\_$hat +/- threshold*Y$\_$hat$\_$se). 

Points are declared outlier if outside this confidence interval. the user choose the **threshold**.

<div style="background-color:rgba(0, 255, 0,0.1); text-align:left; vertical-align: center; padding:10px 0;">
`FuncDetectPointOutlierLocFit`: detection of outlier points in time courses

* @param datain input dataframe. This dataframe contains a set of time courses
* @param myparam character, name of the variable to model in datain (for example, Biomass, PH or LA and so on)
* @param mytime  character, name of the time variable in datain which must be numeric
* @param myid    character, name of the id variable in datain
* @param mylevel numeric, factor to calculate the confidence interval. Increase mylevel to exclude less outliers
* @param mylocfit numeric, The constant component of the smoothing parameter. (see the locfit()) Increase mylocfit to have a very smooth curve

@return a data.frame:

*Ref: the id variable
*mytime: name of the time variable in datain
*myparam: name of the modeled variable in datain
*ypred: the locfit prediction
*sd_ypred: standard deviation of the prediction
*lwr: lower bound of the confidence interval
*upr: upper bound of the confidence interval
*outlier: flag of detected outlier (0 is outlier, 1 is not)

</div>

If a time course has less than 6 points, no smoothing would be done and a warning appears.

```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  library(lubridate)
  library(dplyr)
  library(locfit)
  library(openSilexStatR)
```

# Import of data
In this vignette, we use a toy data set of the *openSilexStatR* library (anonymized real data set). 
```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  mydata<-plant1
  str(mydata)

  mydata<-filter(mydata,!is.na(mydata$thermalTime))
```

# Outlier points detection
I have chosen a smoothing parameter of 30 and a threshold of 10 to detect the outlier points.

```{r,echo=TRUE,message=FALSE, warning=FALSE,error=FALSE}
  # Selection of only 2 genotypes to speed up the process
  plantSel <- c("11430_H","A310_H")
  mydataSub <- filter(mydata,genotypeAlias %in% plantSel)

  resu1<-FuncDetectPointOutlierLocFit(datain=mydataSub,myparam="biovolume",
               mytime="thermalTime",myid="potAlias",
               mylevel=10,
               mylocfit=30)
```


# Graphic

```{r}
plotDetectPointOutlierLocFit(datain=mydataSub,resuin=resu1,
                              myparam="biovolume",
                              mytime="thermalTime",
                              myid="potAlias")
```

# Session info
```{r session,echo=FALSE,message=FALSE, warning=FALSE}
  sessionInfo()
```

# References
1. R Development Core Team (2015). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. ISBN 3-900051-07-0, URL http://www.R-project.org.
2. Catherine Loader (2013). locfit: Local Regression, Likelihood and Density Estimation.. R package version 1.5-9.1. https://CRAN.R-project.org/package=locfit
 
